A Classical Analog of Quantum Interaction-Free Measurement

### 2015-07-17 ms/ifm.tex
ready to submit?

\documentclass[12pt,aps,prb,preprint]{revtex4}
\usepackage{pgf}

@@

There are also counter-intuitive quantum phenonema not involving
entanglement.  Imagine a photon, initially in a ``vertical'' polarized
state.  It is sent through a Faraday rotator, and then through a
vertically polarizing sheet.  The former rotates the polarization
plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where
$N$ is large.  The latter effectively measures the polarization state
in the vertical/horizontal basis; if the result comes out horizontal,
the photon is absorbed, otherwise it goes through in a vertical
eigenstate.  Since $N$ is large, the chances of being absorbed are
small.  More specifically, after the rotation the {\em amplitude\/} to
be in the horizontal state will be $\sim1/N$ so the probability to be
in that mode --- and therefore the probability of absorption --- will
be $\sim1/N^2$.  Thus, most likely, the photon will get slightly
rotated and then reset to vertical.  If the photon is re-routed back
to the apparatus $N$ times, it will still most likely emerge
vertically polarized.  The rotator and the polarizer appear to cancel
each other out.  The general phenomenon is known as the quantum Zeno
effect,\cite{1977JMP....18..756M,PhysRevA.41.2295} but it can also be
described in everyday language, in terms of waking and sleeping
states.\cite{orzel2009teach}

What happens if we now remove the polarizer?  Clearly, the $N$ Faraday
rotations will then convert the photon to a horizontal state.  Thus,
the net effect of the polarizer is to rotate a photon from horizontal
to vertical.  Despite the fact that the polarizer is really just an
absorber of horizontal photons, it can, within this experimental
set-up, have the effect of rotating photons instead of absorbing them.
This is the essence of interaction-free measurement; we can infer, or
`measure' the presence or absence of the polarizer by measuring the
polarization.  If the polarization is rotated we know that the
absorber cannot have been present -- and there is then of course no
absorption -- while if the polarization is not changed then we know
that the absorber was present, but again, at least in the limit $N
\rightarrow \infty$, no absorption actually takes place.

@@

In this paper, we argue that interaction-free measurement need not be
quantum at all.  Rather, the characteristics features of (a)~an
amplitude that cycles between two states, but (b)~an absorber can
block this cycling, thus revealing its presence, but without actually
absorbing any energy, can also appear in classical systems.  If one
has a pair of identical coupled oscillators $x$ and $y$, and $x$ is
set in motion, one will see the strength of the oscillation of $x$
slowly decrease with time, and over longer times the energy will
periodically slosh back and forth between the two systems.  But if
instead oscillator $y$ is damped --- so it is able to {\em absorb\/}
energy --- the rate of decrease of the energy in oscillator $x$ will
be suppressed.  Thus one can infer the ability of oscillator $y$ to
absorb energy by observing the behaviour of oscillator $x$ without, in
the limit that the damping is strong, there actually being any energy
absorbed.

In the other sections of this paper, we flesh out the above argument.
In section~\ref{sec:interf} we work through an alternative well-known
description of interaction-free measurement, more closely resembles
some of the actual experiments.  This is in terms of a Mach-Zehnder
interferometer, whose two arms replace the horizontal and vertical
polarization states of the above thought-experiment.
Figure~\ref{fig:mz} schematically shows the $N=1$ case; the general
case would be a concatenation of such devices.  With an absorber, the
amplitude sloshes between the 'lower' and 'upper' arms of the
interferometer; the absorber results in confining the amplitude to the
lower arm.  Section~\ref{sec:oscill} briefly solves for the
oscillators $x$ and $y$ from the previous paragraph.

\section{An interferometer as an absorber detector}\label{sec:interf}

@@

The contrasting results (\ref{eqn:no-absorber}) and
(\ref{eqn:absorber}) express the characteristic feature of
interaction-free measurement.  With no absorber the amplitude cycles
between $\ket0$ and $\ket1$.  An absorber blocks the sloshing, without
necessarily absorbing.

\section{Coupled oscillators as an absorber detector}\label{sec:oscill}

@@

Consider two coupled oscillators, one of them damped.  Such a system
can be written, with a suitable choice of units, as
\begin{equation}
\begin{array}{l}
\ddot x + x + \kappa y = 0 \\
\ddot y + y + \kappa x + \gamma y = 0 \,.
\label{eqn:coupled}
\end{array}
\end{equation}
Without the damping term ($\gamma=0$), it is easy to verify that
\begin{equation}
\begin{array}{l}
2x = \cos(\sqrt{1+\kappa}\,t) + \cos(\sqrt{1-\kappa}\,t)  \\
2y = \cos(\sqrt{1+\kappa}\,t) - \cos(\sqrt{1-\kappa}\,t)
\end{array}
\end{equation}
is a solution for the initial conditions
\begin{equation}
x(0) = 1, y(0) = \dot x(0) = \dot y(0) = 0 \,.
\label{eqn:inic}
\end{equation}
If the coupling is weak ($\kappa\ll1$) the solution takes the form
\begin{equation}
\begin{array}{l}
x = \cos\left(\frac12\kappa t\right) \cos t \\
y = -\sin\left(\frac12\kappa t\right) \sin t \\
\end{array}
\end{equation}
and the amplitude sloshes from $x$ to $y$.

>>>>

\documentclass[prb,preprint]{revtex4-1}
\usepackage{pgf}

@@

There are also counter-intuitive quantum phenonema not involving
entanglement.  Imagine a photon, initially in a `vertical' polarized
state.  It is sent through a Faraday rotator, and then through a
vertically polarizing sheet.  The former rotates the polarization
plane towards `horizontal' by a small angle, say $\pi/(2N)$, where
$N$ is large.  The latter effectively measures the polarization state
in the vertical/horizontal basis; if the result comes out horizontal,
the photon is absorbed, otherwise it goes through in a vertical
eigenstate.  Since $N$ is large, the chances of being absorbed are
small.  More specifically, after the rotation the {\em amplitude\/} to
be in the horizontal state will be $\sim1/N$ so the probability to be
in that mode --- and therefore the probability of absorption --- will
be $\sim1/N^2$.  Thus, most likely, the photon will get slightly
rotated and then reset to vertical.  If the photon is re-routed back
to the apparatus $N$ times, it will still most likely emerge
vertically polarized.  The rotator and the polarizer appear to cancel
each other out.  The general phenomenon is known as the quantum Zeno
effect,\cite{1977JMP....18..756M,PhysRevA.41.2295} but it can also be
described in everyday language, in terms of waking and sleeping
states.\cite{orzel2009teach}

What happens if we now remove the polarizer?  Clearly, the $N$ Faraday
rotations will then convert the photon to a horizontal state.  Thus,
the net effect of the polarizer is to rotate a photon from horizontal
to vertical.  Despite the fact that the polarizer is really just an
absorber of horizontal photons, it can, within this experimental
set-up, have the effect of rotating photons instead of absorbing them.
This is the essence of interaction-free measurement; we can infer, or
`measure', the presence or absence of the polarizer by measuring the
polarization.  If the polarization is rotated we know that the
absorber cannot have been present -- and there is then of course no
absorption -- while if the polarization is not changed then we know
that the absorber was present, but again, at least in the limit $N
\rightarrow \infty$, no absorption actually takes place.

@@

In this paper, we argue that interaction-free measurement need not be
quantum at all.  Rather, the characteristic features of (a)~an
amplitude that cycles between two states, but (b)~an absorber can
block this cycling, thus revealing its presence, but without actually
absorbing any energy, can also appear in classical systems.  If one
has a pair of identical coupled oscillators $x$ and $y$, and $x$ is
set in motion, one will see the strength of the oscillation of $x$
slowly decrease with time, and over longer times the energy will
periodically slosh back and forth between the two systems.  But if
instead oscillator $y$ is damped --- so it is able to {\em absorb\/}
energy --- the rate of decrease of the energy in oscillator $x$ will
be suppressed.  Thus one can infer the ability of oscillator $y$ to
absorb energy by observing the behaviour of oscillator $x$ without, in
the limit that the damping is strong, there actually being any energy
absorbed.

In the other sections of this paper, we flesh out the above argument.
In section~\ref{sec:interf} we work through an alternative well-known
description of interaction-free measurement, more closely resembles
some of the actual experiments.  This is in terms of a Mach-Zehnder
interferometer, whose two arms replace the horizontal and vertical
polarization states of the above thought-experiment.
Figure~\ref{fig:mz} schematically shows the $N=1$ case; the general
case would be a concatenation of such devices.  With an absorber, the
amplitude sloshes between the `lower' and `upper' arms of the
interferometer; the absorber results in confining the amplitude to the
lower arm.  In section~\ref{sec:oscill} we validate the foregoing
argument by briefly solving for the coupled oscillators $x$ and $y$
from the previous paragraph.

\section{Interferometers as detectors of absorbers}\label{sec:interf}

@@

The contrasting results (\ref{eqn:no-absorber}) and
(\ref{eqn:absorber}) express the characteristic feature of
interaction-free measurement.  With no absorber, the amplitude out of
the $N$-th beam splitter sloshes between $\ket0$ and $\ket1$ as $N$
increases.  An absorber blocks the sloshing, without necessarily
absorbing.

\section{Coupled oscillators as detectors of absorbers}\label{sec:oscill}

@@

Consider two coupled oscillators, $x,y$, of which $y$ is damped.  Such
a system can be written, with a suitable choice of units, as
\begin{equation}
\begin{array}{l}
\ddot x + x + \kappa y = 0 \\
\ddot y + y + \kappa x + \gamma y = 0 \,.
\label{eqn:coupled}
\end{array}
\end{equation}
Without the damping term ($\gamma=0$), it is easy to verify that
\begin{equation}
\begin{array}{l}
2x = \cos(\sqrt{1+\kappa}\,t) + \cos(\sqrt{1-\kappa}\,t)  \\
2y = \cos(\sqrt{1+\kappa}\,t) - \cos(\sqrt{1-\kappa}\,t)
\end{array}
\end{equation}
is a solution for the initial conditions
\begin{equation}
x(0) = 1, y(0) = \dot x(0) = \dot y(0) = 0 \,.
\label{eqn:inic}
\end{equation}
If the coupling is weak ($\kappa\ll1$) the solution takes the form
\begin{equation}
\begin{array}{l}
x = \cos\left(\frac12\kappa t\right) \cos t \\
y = -\sin\left(\frac12\kappa t\right) \sin t \\
\end{array}
\end{equation}
and the amplitude sloshes from $x$ to $y$.
